---
title: "Storm Damage Analysis: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
  pdf_document:
    toc: true
---

# Storm Damage Analysis: Peer Assessment 2

## Abstract

This analysis decomposes weather event data from NOAA and provides an overview of the most frequent
causes of fatalities, injuries, crop damage, and property damage.  A summary of the most significant
types of events is provided.  Furthermore, the fatalities/injuries and crop/property damage are
compared by region of the country.  In most regions of the country, deaths are frequently caused by
excessive heat.  Injuries are frequently caused by tornadoes.  Significant crop damage is caused by
hail and floods, and significant property damage is caused by tornadoes.

## Data Processing

The data for this analysis is obtained from the U.S. National Oceanic and Atmospheric
Administration's (NOAA) storm database.  An archived version of this data can be found from the
Reproducible Research course stored in CloudFront.net.

```{r, cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
              "data/stormdata.csv.bz2", method="curl")
stormData = read.csv(bzfile("data/stormdata.csv.bz2"))
```

Build up a method of categorizing states by region for further data analysis.

```{r, cache=TRUE}
regionStatesNorthEast = c(
    "ME", "VT", "NH", "MA", "CT", "RI",
    "NY", "PA", "NJ");
regionStatesMidWest = c(
    "ND", "SD", "NE", "KS", "MN", "IA", "MO",
    "WI", "IL", "MI", "IN", "OH");
regionStatesSouth = c(
    "WV", "MD", "DE", "DC", "VA", "NC", "SC", "GA", "FL",
    "KY", "TN", "MS", "AL",
    "OK", "TX", "AR", "LA");
regionStatesWest = c(
    "WA", "OR", "CA", "AK",
    "MT", "ID", "WY", "NV", "UT", "CO", "AZ", "NM");

# An associative array mapping state to region.
regionByState = c(
    rep("northeast", length(regionStatesNorthEast)),
    rep("midwest", length(regionStatesMidWest)),
    rep("south", length(regionStatesSouth)),
    rep("west", length(regionStatesWest)))
names(regionByState) = c(
         regionStatesNorthEast,
         regionStatesMidWest,
         regionStatesSouth,
         regionStatesWest)

# A function to convert state to region or NA.
getRegionForState = function(state) {
    stateName = as.character(state);
    if (stateName %in% names(regionByState)) {
        return(regionByState[[stateName]]);
    } else {
        return(NA);
    }
}

# Finally define a field REGION for each observation.
stormData$REGION = as.factor(sapply(stormData$STATE, getRegionForState))
```

## Results

### Health Damage

*Across the United States, which types of events (as indicated in the EVTYPE variable) are most
harmful with respect to population health?*

There are over 985 types of events, we need to limit ourselves to only the most dangerous ones.
The top 20 most dangerous types of events can be seen below:
```{r}
injuriesByEvType = with(stormData, aggregate(INJURIES, by=list(EVTYPE), sum))
names(injuriesByEvType) = c("EVTYPE", "INJURIES")
injuriesByEvType = injuriesByEvType[order(injuriesByEvType$INJURIES, decreasing=TRUE), ]

head(injuriesByEvType, 20)
```

The most fatal event types can be seen as well.
```{r}
fatalitiesByEvType = with(stormData, aggregate(FATALITIES, by=list(EVTYPE), sum))
names(fatalitiesByEvType) = c("EVTYPE", "FATALITIES")
fatalitiesByEvType = fatalitiesByEvType[order(fatalitiesByEvType$FATALITIES, decreasing=TRUE), ]

head(fatalitiesByEvType, 20)
```

We can further break down the injuries and fatalities by region.  The following graphs show the 5
most inuring or fatal events broken down by region.

```{r}
fatalitiesByEventAndRegion = with(
    stormData[stormData$EVTYPE %in% head(fatalitiesByEvType$EVTYPE, 5), ],
    aggregate(FATALITIES, by=list(EVTYPE, REGION), sum))
names(fatalitiesByEventAndRegion) = c("EVTYPE", "REGION", "FATALITIES")

injuriesByEventAndRegion = with(
    stormData[stormData$EVTYPE %in% head(injuriesByEvType$EVTYPE, 5), ],
    aggregate(INJURIES, by=list(EVTYPE, REGION), sum))
names(injuriesByEventAndRegion) = c("EVTYPE", "REGION", "INJURIES")

library(ggplot2)
library(gridExtra)

fatalitiesPlot =
    ggplot(fatalitiesByEventAndRegion, aes(x=REGION, y=FATALITIES)) +
    ggtitle("Most Fatal Events") +
    facet_grid(EVTYPE ~ .) +
    geom_point()
injuriesPlot =
    ggplot(injuriesByEventAndRegion, aes(x=REGION, y=INJURIES)) +
    ggtitle("Most Injurious Events") +
    facet_grid(EVTYPE ~ .) +
    geom_point()
grid.arrange(fatalitiesPlot, injuriesPlot, ncol=2)
```

### Economic Damage

*Across the United States, which types of events have the greatest economic consequences?*

Like we did with fatalities and injuries, we need to limit the number of types of events we consider.

The top 20 most damaging types of events to crops can be seen below:

```{r}
cropDmgByEvType = with(stormData, aggregate(CROPDMG, by=list(EVTYPE), sum))
names(cropDmgByEvType) = c("EVTYPE", "CROPDMG")
cropDmgByEvType = cropDmgByEvType[order(cropDmgByEvType$CROPDMG, decreasing=TRUE), ]

head(cropDmgByEvType, 20)
```

The top 20 most damaging types of events to property can be seen below:

```{r}
propDmgByEvType = with(stormData, aggregate(PROPDMG, by=list(EVTYPE), sum))
names(propDmgByEvType) = c("EVTYPE", "PROPDMG")
propDmgByEvType = propDmgByEvType[order(propDmgByEvType$PROPDMG, decreasing=TRUE), ]

head(propDmgByEvType, 20)
```

We can further break down the crop and propery damage by region.  The following graphs show the 5
most damaging events broken down by region.

```{r}
cropDmgByEventAndRegion = with(
    stormData[stormData$EVTYPE %in% head(cropDmgByEvType$EVTYPE, 5), ],
    aggregate(CROPDMG, by=list(EVTYPE, REGION), sum))
names(cropDmgByEventAndRegion) = c("EVTYPE", "REGION", "CROPDMG")

propDmgByEventAndRegion = with(
    stormData[stormData$EVTYPE %in% head(propDmgByEvType$EVTYPE, 5), ],
    aggregate(PROPDMG, by=list(EVTYPE, REGION), sum))
names(propDmgByEventAndRegion) = c("EVTYPE", "REGION", "PROPDMG")

library(ggplot2)
library(gridExtra)

cropDmgPlot =
    ggplot(cropDmgByEventAndRegion, aes(x=REGION, y=CROPDMG)) +
    ggtitle("Most Crop Damaging Events") +
    facet_grid(EVTYPE ~ .) +
    geom_point()
propDmgPlot =
    ggplot(propDmgByEventAndRegion, aes(x=REGION, y=PROPDMG)) +
    ggtitle("Most Property Damaging Events") +
    facet_grid(EVTYPE ~ .) +
    geom_point()
grid.arrange(cropDmgPlot, propDmgPlot, ncol=2)
```

## Conclusions

All regions of the country face a significant number of deaths from Excessive Heat, with the West
being a notable exception.  The Midwest in particular has a high rate of deaths from Heat and
Tornados.

The Midwest and the South suffer a high rate of injury from Tornadoes.

The Midwest and the South suffer high amounts of crop damage from Hail, Flash Floods, and Floods.

The Midwest and the South suffer high amounts of property damage from Floods, Tornadoes, and
Thunderstorm Winds.
